---
title: "00-import-BioTIME-database"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: github_document
params:
  name: "00-import-BioTIME-database" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "92e0b108-b72d-4d77-bf7a-4e46e4cde787")
```

The purpose of this document is to import and have a first look into the BioTIME database from *BioTIME: A database of biodiversity time series for the Anthropocene*.

```{r packages, meassage = F}
library("conflicted")
library(dplyr)
library(data.table)
library(ggplot2)
library(stringr)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```


## Data

#### Read files

```{r}
#### read the file
dt_fullquery <- 
  fread("input_data/BioTIME/BioTIMEquery_24_06_2021.csv")

colnames(dt_fullquery)[1] <- "ID_SAMPLE"

# remove whitespaces at beginning/end of species name
dt_fullquery[, GENUS_SPECIES := str_trim(GENUS_SPECIES, side = c("both"))] %>% 
  # correct spelling mistake
  .[GENUS == "Zanthoxylem", GENUS := "Zanthoxylum"]

# number of studies
print(paste("number of studies:", 
            dt_fullquery$STUDY_ID %>% uniqueN()))
```

#### Prepare for phyloseq

```{r}
# show list of duplicated species IDs which will be removed in tax table
dt_fullquery[,.(ID_SPECIES, GENUS, SPECIES)] %>% 
  unique() %>%
  .[duplicated(ID_SPECIES)]
    
# create taxonomic table
dt_tax_table <-
  dt_fullquery[,.(ID_SPECIES, GENUS, SPECIES)] %>% 
  unique() %>%
  .[!duplicated(ID_SPECIES)]

# extract sample information
dt_sample_info <-
  dt_fullquery[,.(ID_SAMPLE, STUDY_ID, DAY, MONTH, YEAR, SAMPLE_DESC, PLOT, 
                  ID_SPECIES, LATITUDE, LONGITUDE)]

# extract count table
dt_count_table <-
  dt_fullquery[,.(ID_SAMPLE, ID_SPECIES,
                  ABUNDANCE = sum.allrawdata.ABUNDANCE)]

```



#### Focus on only one study (ID = 63)

```{r}
# focus on one study
study_id = 63
```


```{r}
# extract one study
dt_query <- 
  dt_fullquery[STUDY_ID == study_id]

knitr::kable(head(dt_query))
```



```{r, include=FALSE}
#### Some code from BioTIMEInteractions_02_04_2018.Rmd


# Maria Dornelas 09.06.2015
rarefysamples <- function(Year, SampleID, Species, Abundance, resamps) {
  #######################################################################
  # takes as input a  Year, SampleID, Species, Abundance and number of resamples
  # which should be in dataframe so that elements match
  # calculates turnover:
  # 1) between each year and the first year
  # 2) between pairs of adjacent years
  # 3) between each year and the lasy year of the time series
  # for the rarefied pooled samples
  ###########################################################################
  
  rareftab <- data.frame(array(NA, dim = c(0, 3)))
  # getting vector with number of samples per year
  nsamples <- c()
  for (y in unique(Year)) {
    nsamples <- c(nsamples, length(unique(SampleID[Year == y])))
  }
  t <- 1
  minsample <- min(nsamples)
  for (repeats in 1:resamps) {
    raref <- data.frame(array(NA, dim = c(1, 3)))
    names(raref) <- c("Year", "Species", "Abundance")
    for (y in unique(Year)) {
      #getting samples for this year
      samps <- unique(SampleID[Year == y])
      # re-sampling samples to equalize number of samples
      sam <- as.character(sample(samps, minsample, replace = T))
      # getting data that belongs to bootstraped samples
      rarefyear <-
        data.frame(SampleID[which(SampleID %in% sam & Year == y)],
                   Species[which(SampleID %in% sam & Year == y)],
                   Abundance[which(SampleID %in% sam & Year == y)])
      names(rarefyear) <- c("SampleID", "Species", "Abundance")
      # calculating pooled abundances of each species to store
      spabun <-
        tapply(as.numeric(rarefyear[, 3]), as.character(rarefyear[, 2]), sum)
      spar <-
        data.frame(rep(y, length(spabun)), names(spabun), spabun, row.names = NULL)
      names(spar) <- c("Year", "Species", "Abundance")
      raref <- rbind(raref, spar)
    }
    # calculating year by species table of abundance
    rareftab <-
      rbind(rareftab, cbind(rep(repeats, dim(raref)[1]), raref))
  }
  return (rareftab)
}
```


```{r, include = FALSE}
TSrf <- list()
IDs <- unique(dt_query$STUDY_ID)

for(i in 1:length(IDs)) {
  data <- dt_query[dt_query$STUDY_ID == IDs[i], ]
  TSrf[[i]] <-
    rarefysamples(
      data$YEAR,
      data$SAMPLE_DESC,
      data$GENUS_SPECIES,
      data$sum.allrawdata.ABUNDANCE,
      1
    )
}
names(TSrf) <- IDs

rf <- do.call(rbind, TSrf)
rf <-
  data.frame(rf, ID = rep(names(TSrf), times = unlist(lapply(TSrf, nrow))))
rf <- rf[!is.na(rf$Year), -1]

#### prepare the rarefied output for diversity metric code
t1 <- with(rf, tapply(Abundance, list(Year, Species, ID), sum))
t2 <- unique(rf[, c("ID", "Year")])

#### produces a list of matrices for each study - in this case is only a single dataset
dsList <- list()

for (i in 1:dim(t1)[3]) {
  id <- dimnames(t1)[[3]][i]
  a <- subset(t2, ID == id)$Year
  b <- t1[dimnames(t1)[[1]] %in% as.character(a), , i]
  dsList[[i]] <- data.frame(Year = rownames(b), b)
}

names(dsList) <- dimnames(t1)[[3]]

#### replacing NA with zero
for(i in 1:(length(dsList))) {
  dsList[[i]][is.na(dsList[[i]])] <- 0
}

write.csv(rf,
          path_target(paste0("BioTIME-Study-", study_id, ".csv")))
```


### plot time series for selected study

```{r, fig.width=10}
  
# ## group species
# rf_grouped <- as.data.table(rf) %>% 
#   # .[, c("Genus", "Species") := tstrsplit(Species, " ")] %>% 
#   .[, .(Abundance_sum = sum(Abundance)),
#         by = c("Year", "Species")]

ggplot(rf, aes(Year, Abundance, col = Species)) +
  geom_line() #+
  # theme(legend.position = "none")

```


### values of selected time series

```{r}
rf_time <- as.data.table(rf) %>% 
  # .[, c("Genus", "Species") := tstrsplit(Species, " ")] %>% 
  dcast(., Species ~ Year, value.var = "Abundance")

knitr::kable(rf_time)
```


<br>

## BioTime Metadata

#### Read file

```{r}
#### read the file
dt_biotimeMeta <-
  fread("input_data/BioTIME/biotimeMetadata_24_06_2021.csv")

```


#### List of all columns

```{r, results = "asis", echo = FALSE}
# list all column names of Metadata
cat(paste("*", colnames(dt_biotimeMeta)), sep="\n")

```


#### First rows of table (only including important columns)

Select some important columns to get a better overview which studies are included in the dataset.

```{r, include=FALSE}
dt_biotimeMeta_small <-
  dt_biotimeMeta[, .(
    STUDY_ID,
    REALM,
    CLIMATE,
    # GENERAL_TREAT,
    # TREATMENT,
    # TREAT_COMMENTS,
    # TREAT_DATE,
    # HABITAT,
    # PROTECTED_AREA,
    # BIOME_MAP,
    TAXA,
    # ORGANISMS,
    TITLE,
    AB_BIO,
    # HAS_PLOT,
    DATA_POINTS,
    START_YEAR,
    END_YEAR,
    # CENT_LAT,
    # CENT_LONG,
    NUMBER_OF_SPECIES,
    NUMBER_OF_SAMPLES,
    NUMBER_LAT_LONG,
    TOTAL,
    # GRAIN_SIZE_TEXT,
    # GRAIN_SQ_KM,
    # AREA_SQ_KM,
    # CONTACT_1,
    # CONTACT_2,
    # CONT_1_MAIL,
    # CONT_2_MAIL,
    # LICENSE,
    # WEB_LINK,
    # DATA_SOURCE,
    # METHODS,
    # SUMMARY_METHODS,
    # LINK_ID,
    # COMMENTS,
    # DATE_STUDY_ADDED,
    ABUNDANCE_TYPE,
    BIOMASS_TYPE
    # SAMPLE_DESC_NAME
  )]

write.csv(dt_biotimeMeta_small,
          path_target("BioTIME_Meta_reduced.csv"),
          row.names = FALSE)

dt_biotimeMeta_small <-
  dt_biotimeMeta_small %>%
  .[, .(
    STUDY_ID,
    REALM,
    CLIMATE,
    TAXA,
    AB_BIO,
    DATA_POINTS,
    NUMBER_OF_SPECIES,
    NUMBER_OF_SAMPLES,
    NUMBER_LAT_LONG,
    TOTAL,
    ABUNDANCE_TYPE,
    BIOMASS_TYPE
  )] %>%
  # take only studys with at least 10 time points (DATA_POINTS)
  .[DATA_POINTS >= 10]
```

Some of the Studies in the BioTIME data are listed in the table below (only studies with at least 10 datapoints):

```{r}
knitr::kable(dt_biotimeMeta_small %>% head(n = 16))
```

### Detailed Info about some columns

##### DATA_POINTS

Number of years covered by study

```{r}
ggplot(dt_biotimeMeta, aes(DATA_POINTS)) +
  geom_histogram(binwidth = 5) +
  scale_x_continuous(breaks = seq(0,100,10))
```


##### AB_BIO

Does study contain abundance, biomass or both (A, B or AB)

```{r}
knitr::kable(dt_biotimeMeta$AB_BIO %>% table(useNA = "ifany"))
```


##### ABUNDANCE_TYPE

```{r}
knitr::kable(dt_biotimeMeta$ABUNDANCE_TYPE %>% table(useNA = "ifany"))
```


##### BIOMASS_TYPE

```{r}
knitr::kable(dt_biotimeMeta$BIOMASS_TYPE %>% table(useNA = "ifany"))

```


##### REALM

```{r}
dt_biotimeMeta$REALM %>% unique()
```

##### CLIMATE

```{r}
dt_biotimeMeta$CLIMATE %>% unique()

```


##### TAXA

```{r}
dt_biotimeMeta$TAXA %>% unique()
```


##### NUMBER_OF_SPECIES

Number of species in study

```{r}
range(dt_biotimeMeta$NUMBER_OF_SPECIES)

ggplot(dt_biotimeMeta, aes(NUMBER_OF_SPECIES)) +
  geom_histogram(binwidth = 100)

ggplot(dt_biotimeMeta[NUMBER_OF_SPECIES < 200], aes(NUMBER_OF_SPECIES)) +
  geom_histogram(binwidth = 1)

```


<!-- ##### NUMBER_OF_SAMPLES -->

<!-- Number of unique samples in study -->

<!-- ```{r} -->
<!-- dt_biotimeMeta$NUMBER_OF_SAMPLES %>% range() -->

<!-- # ggplot(dt_biotimeMeta, aes(NUMBER_OF_SAMPLES)) + -->
<!-- #   geom_histogram(binwidth = 100) -->

<!-- ggplot(dt_biotimeMeta[NUMBER_OF_SAMPLES < 100], aes(NUMBER_OF_SAMPLES)) + -->
<!--   geom_histogram(binwidth = 1) -->

<!-- ``` -->


##### NUMBER_LAT_LONG

Number of studies at **1 location** (NUMBER_LAT_LONG == 1):
**`r dt_biotimeMeta[NUMBER_LAT_LONG == 1, .N]`**

Number of studies at **multiple locations** (NUMBER_LAT_LONG > 1): 
**`r dt_biotimeMeta[NUMBER_LAT_LONG > 1, .N]`**

Max. NUMBER_LAT_LONG: **`r dt_biotimeMeta$NUMBER_LAT_LONG %>% max()`**




## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
