---
title: "Different time series"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: github_document
number_sections: true
params:
  name: "01-different-time-series" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "964cf30c-e7ec-4fee-a704-f01c350f8766")
```

The purpose of this document is to give an overview of different time series that can be used in the further process.
These time series can differ in

  * number of species
  * number of time points

#### Packages

```{r packages}
library("conflicted")
library(dplyr)
library(data.table)
library(ggplot2)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## BioTIME

#### Read Data

```{r}
#### read the file
dt_fullquery <- 
  fread("data/BioTIME/BioTIMEquery_24_06_2021.csv")

#### read BioTIME Metadata
dt_biotimeMeta <-
  fread("data/00-import-BioTIME-database/BioTIME_Meta_reduced.csv")

```


#### Criteria to select a suitable study

 * there should be enough DATA_POINTS
 * certain Abundance Type ("Count")

```{r}
# Example
dt_biotimeMeta[DATA_POINTS > 10 & ABUNDANCE_TYPE == "Count" & NUMBER_LAT_LONG == 1 & NUMBER_OF_SPECIES > 50]$STUDY_ID
```



```{r, include=FALSE}
# function to extract one study out of the full query with adjusted Date column
extract_study <- function(data, ID) {
  
  tmp <- data[STUDY_ID == ID]
  
  # check if whole dates are given or only years
  if(tmp[!is.na(DAY) | !is.na(MONTH), .N] != 0) {
    tmp[, Date := as.Date(paste0(YEAR, "-", MONTH, "-", DAY), format = "%Y-%m-%d")]
  } else {
    tmp[, Date := YEAR]
  }
  
  tmp[, .(Date, Abundance = sum.allrawdata.ABUNDANCE,
          Species = SPECIES, Genus = GENUS, GENUS_SPECIES = GENUS_SPECIES)]
}
```


## Bad Example (with gaps in time)

```{r}
study_id = 152

# extract data of the specific study
dt_study <-
  extract_study(dt_fullquery, ID = study_id)

# summarize over multiple entries in the same year
dt_study_grouped <-
  dt_study[, .(Abundance_mean = mean(Abundance)),
    by = c("Date", "Genus")]


# plot timeseries
ggplot(dt_study, aes(Date, Abundance)) +
  geom_line(aes(col = GENUS_SPECIES))  +
  theme(legend.position = "none")

# plot Abundance summed up for same Date
ggplot(dt_study_grouped, aes(Date, Abundance_mean)) +
  geom_line(aes(col = Genus))  +
  theme(legend.position = "none")

# plot only first part
ggplot(dt_study_grouped[Date < "1990-01-01"], aes(Date, Abundance_mean)) +
  geom_line(aes(col = Genus)) 

```

```{r}
dt_study_grouped[Genus == "Fritillaria"] %>% View()

# multiple values on consecutive days (e.g. 1982-06-XY)
```



## 2nd Example

Example with multiple locations.

```{r}
study_id = 163
```

#### Information on the study

```{r, results = "asis", echo = FALSE}
study_Meta <-
  dt_biotimeMeta[STUDY_ID == study_id]

table_Meta <-
  data.table(Column = colnames(study_Meta),
             Info = as.character(study_Meta[1,]))

knitr::kable(table_Meta)

```

### time series plots

```{r}

# extract data of the specific study
dt_study <-
  extract_study(dt_fullquery, ID = study_id)

# summarize over multiple entries in the same year
dt_study_grouped <-
  dt_study[, .(Abundance_mean = mean(Abundance)),
    by = c("Date", "Genus")]


# plot timeseries
ggplot(dt_study, aes(Date, Abundance)) +
  geom_line(aes(col = GENUS_SPECIES))  +
  theme(legend.position = "none")

# plot Abundance summed up for same Date
ggplot(dt_study_grouped, aes(Date, Abundance_mean)) +
  geom_line(aes(col = Genus))  +
  theme(legend.position = "none")

# same plot with y log-scaled
ggplot(dt_study_grouped, aes(Date, Abundance_mean)) +
  geom_line(aes(col = Genus))  +
  theme(legend.position = "none") +
  scale_y_log10()

dt_study_grouped[Abundance_mean < 10]
```




## 3rd Example

```{r}
study_id = 225
```

#### Information on the study

```{r, results = "asis", echo = FALSE}
study_Meta <-
  dt_biotimeMeta[STUDY_ID == study_id]

table_Meta <-
  data.table(Column = colnames(study_Meta),
             Info = as.character(study_Meta[1,]))

knitr::kable(table_Meta)

```

### time series plots

```{r}

# extract data of the specific study
dt_study <-
  extract_study(dt_fullquery, ID = study_id)

# summarize over multiple entries in the same year
dt_study_grouped <-
  dt_study[, .(Abundance_mean = mean(Abundance)),
    by = c("Date", "Genus")]


# plot timeseries
ggplot(dt_study, aes(Date, Abundance)) +
  geom_line(aes(col = GENUS_SPECIES))  +
  theme(legend.position = "none")

# plot Abundance summed up for same Date
ggplot(dt_study_grouped, aes(Date, Abundance_mean)) +
  geom_line(aes(col = Genus))  +
  theme(legend.position = "none")

# same plot with y log-scaled
ggplot(dt_study_grouped, aes(Date, Abundance_mean)) +
  geom_line(aes(col = Genus))  +
  theme(legend.position = "none") +
  scale_y_log10()

```


## 4th Example

```{r}
study_id = 57
```

#### Information on the study

```{r, results = "asis", echo = FALSE}
study_Meta <-
  dt_biotimeMeta[STUDY_ID == study_id]

table_Meta <-
  data.table(Column = colnames(study_Meta),
             Info = as.character(study_Meta[1,]))

knitr::kable(table_Meta)

```

### time series plots

```{r}

# extract data of the specific study
dt_study <-
  extract_study(dt_fullquery, ID = study_id)

# summarize over multiple entries in the same year
dt_study_grouped <-
  dt_study[, .(Abundance_mean = mean(Abundance)),
    by = c("Date", "Genus")]

# number of entries per Date and Genus
dt_study[, .N, by= c("Date", "Genus")]


# plot timeseries
ggplot(dt_study, aes(Date, Abundance)) +
  geom_line(aes(col = GENUS_SPECIES))  +
  theme(legend.position = "none")

# plot Abundance summed up for same Date
ggplot(dt_study_grouped, aes(Date, Abundance_mean)) +
  geom_line(aes(col = Genus))  +
  theme(legend.position = "none")

# same plot with y log-scaled
ggplot(dt_study_grouped, aes(Date, Abundance_mean)) +
  geom_line(aes(col = Genus))  +
  theme(legend.position = "bottom") +
  scale_y_log10()

```




## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
