---
title: "Time series (BioTIME)"
Time: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: github_document
number_sections: true
params:
  name: "01a-timeseries-BioTIME" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "964cf30c-e7ec-4fee-a704-f01c350f8766")
```

The purpose of this document is to analyse different time series from the BioTIME database that could be interesting for the further process.

#### Packages

```{r packages}
library("conflicted")
library(dplyr)
library(data.table)
library(ggplot2)
library(viridis) # for color palettes
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## BioTIME

#### Read Data

```{r}
#### read the file
dt_fullquery <- 
  fread("data/BioTIME/BioTIMEquery_24_06_2021.csv")

#### read BioTIME Metadata
dt_biotimeMeta <-
  fread("data/00-import-BioTIME-database/BioTIME_Meta_reduced.csv")

```


#### Criteria to select a suitable study

 * there should be enough DATA_POINTS (>50)
 * certain Abundance Type ("Count") --> not so important
 * the samples should be taken from only one location
 * the intervals between the time steps should be equal (e.g. one year)

```{r}
# show available studies
dt_biotimeMeta[DATA_POINTS > 35 & NUMBER_LAT_LONG == 1] %>%
# dt_biotimeMeta[DATA_POINTS > 35 & NUMBER_LAT_LONG == 1 & AB_BIO %in% c("AB", "A")] %>%
  .[, YEARS := paste0(START_YEAR, "-", END_YEAR)] %>% 
  .[, c("TITLE", "AB_BIO", "REALM", "CLIMATE", "START_YEAR", "END_YEAR") := NULL]%>% 
  .[order(-DATA_POINTS)] %>% 
  knitr::kable()
```



```{r, include=FALSE}
# function to extract one study out of the full query with adjusted Time column
extract_study <- function(data, ID) {
  
  tmp <- data[STUDY_ID == ID]
  
  # check if whole Times are given or only years
  if(tmp[!is.na(DAY) | !is.na(MONTH), .N] != 0) {
    tmp[, Time := as.Date(paste0(YEAR, "-", MONTH, "-", DAY), format = "%Y-%m-%d")]
  } else {
    tmp[, Time := YEAR]
  }
  
  tmp[, .(Time, Abundance = sum.allrawdata.ABUNDANCE,
          Species = GENUS_SPECIES, Genus = GENUS)]
}
```




We need Time Series with at least 50 time steps.

Further Criteria for useful time series:

 * >50 time points
 * time steps should be at equal intervals
 * location of sampling should be only one
 
## Data Preprocessing

Function to preprocess data given a study ID and plot abundances afterwards.

```{r}

study_processing_bioTIME <-
  function(study_id, th_species = 0.05, th_genus = 0.05) {
    
    # Metadata (Information about this study)
    study_Meta <-
      dt_biotimeMeta[STUDY_ID == study_id]
    
    table_Meta <-
      data.table(Column = colnames(study_Meta),
                 Info = as.character(study_Meta[1,]))
    
    knitr::kable(table_Meta)
    
    
    # 1)  extract data of the specific study
    dt_study <-
      extract_study(dt_fullquery, ID = study_id)
    
    ## add zeros for missing values
    
    ### generate data.table with all combinations of Species/Genus and Time
    all_combinations <-
      as.data.table(tidyr::crossing(unique(dt_study$Time), unique(dt_study$Species)))
    colnames(all_combinations) <- c("Time", "Species")
    all_combinations[, Abundance := 0] %>% 
      .[, c("Genus", "tmp") := tstrsplit(Species, " ")] %>% 
      .[, tmp := NULL]
    
    ### merge combinations and zero values to dt_study
    dt_study <- 
      merge(dt_study, all_combinations,
            by = c("Time", "Species", "Genus"),
            all = T) %>% 
      .[, Abundance := Abundance.x] %>% 
      .[is.na(Abundance.x), Abundance := Abundance.y] %>% 
      .[, c("Abundance.x", "Abundance.y") := NULL]
    
    
    # 2)  summarize over multiple entries in the same year
    dt_study_genus <-
      dt_study[, .(Abundance_sum = sum(Abundance)),
        by = c("Time", "Genus")]
    
    
    # 3)  add relative abundance
    
    ## Species
    dt_study[, Abundance_rel := Abundance/ sum(Abundance), 
             by = c("Time")]
    
    ## Genus
    dt_study_genus[, Abundance_rel := Abundance_sum / sum(Abundance_sum), 
             by = c("Time")]
    
    
    # 3)  Group Species/Genus with relative abundance below a certain threshold as "other"
    
    ## add Species "other" for max. rel. abundance < threshold
    Species_other <-
      dt_study[, .(max_Abundance = max(Abundance_rel)),
             by = c("Species")] %>% 
      .[max_Abundance < th_species, .(Species)]
    
    dt_study_species <-
      dt_study[, Species_grouped := Species] %>% 
      .[Species %in% Species_other$Species, Species_grouped := "other"]
    
    ## add Genus "other" for max. rel. abundance < threshold
    Genus_other <-
      dt_study_genus[, .(max_Abundance = max(Abundance_rel)),
             by = c("Genus")] %>% 
      .[max_Abundance < th_genus, .(Genus)]
    
    dt_study_genus[, Genus_grouped := Genus] %>% 
      .[Genus %in% Genus_other$Genus, Genus_grouped := "other"]
    
    
    # 5)  sum up abundances for group "other"
    
    ## Species
    dt_study_species <-
      dt_study_species[, .(Abundance_rel = sum(Abundance_rel)),
        by = c("Time", "Species_grouped")]
    ## Genus
    dt_study_genus <-
      dt_study_genus[, .(Abundance_sum = sum(Abundance_sum),
                         Abundance_rel = sum(Abundance_rel)),
        by = c("Time", "Genus_grouped")]
    
    
    ### plot timeseries
    
    # by Species, total
    pl_species_total <-
      ggplot(dt_study, aes(Time, Abundance)) +
      geom_line(aes(col = Species))  +
      labs(title = "by Species, total Abundances") +
      theme(legend.position = "right")
    
    # by Species, relative
    pl_species_rel <-
      ggplot(dt_study_species, aes(Time, Abundance_rel)) +
      geom_bar(aes(fill = Species_grouped), stat = "identity") +
      labs(title = "by Species, relative Abundances") +
      theme(legend.position = "right") +
      scale_fill_viridis(discrete = TRUE, option = "turbo") 
    
    # by Genus, total (Abundance summed up)
    pl_genus_total <-
      ggplot(dt_study_genus, aes(Time, Abundance_sum)) +
      geom_line(aes(col = Genus_grouped))  +
      labs(title = "by Genus, total Abundances") +
      theme(legend.position = "right") +
      scale_color_viridis(discrete = TRUE, option = "turbo") 
    
    # by Genus, relative (Abundance summed up)
    pl_genus_rel <-
      ggplot(dt_study_genus, aes(Time, Abundance_rel)) +
      geom_bar(aes(fill = Genus_grouped), stat = "identity")  +
      labs(title = "by Genus, relative Abundances") +
      theme(legend.position = "right") +
      scale_fill_viridis(discrete = TRUE, option = "turbo") 
    
    print(pl_species_total)
    print(pl_species_rel)
    print(pl_genus_total)
    print(pl_genus_rel)
  }

```


# Examples

## (1)

```{r}
study_id = 339

study_processing_bioTIME(study_id)
```

<!-- #### Information on the study -->

<!-- ```{r, results = "asis", echo = FALSE} -->
<!-- study_Meta <- -->
<!--   dt_biotimeMeta[STUDY_ID == study_id] -->

<!-- table_Meta <- -->
<!--   data.table(Column = colnames(study_Meta), -->
<!--              Info = as.character(study_Meta[1,])) -->

<!-- knitr::kable(table_Meta) -->

<!-- ``` -->


<!-- ### extraction and preparation of the time series -->

<!-- Preprocessing steps of the sample data for each study: -->

<!-- 1. extracting the study ID from the whole dataset -->
<!-- 2. for Genus add Abundances at the same Date / same Genus -->
<!-- 3. calculating relative abundance by -->
<!--     a) Species -->
<!--     b) Genus (sum of abundance over the samples) -->
<!-- 4. Check how many Species/Genus have relative abundance below certain threshold -->
<!--    -> group them as "other" -->
<!-- 5. sum up abundances for group "other" -->


<!-- ```{r} -->
<!-- # 1)  extract data of the specific study -->
<!-- dt_study <- -->
<!--   extract_study(dt_fullquery, ID = study_id) -->

<!-- ## add zeros for missing values -->

<!-- ### generate data.table with all combinations of Species/Genus and Time -->
<!-- all_combinations <- -->
<!--   as.data.table(tidyr::crossing(unique(dt_study$Time), unique(dt_study$Species))) -->
<!-- colnames(all_combinations) <- c("Time", "Species") -->
<!-- all_combinations[, Abundance := 0] %>%  -->
<!--   .[, c("Genus", "tmp") := tstrsplit(Species, " ")] %>%  -->
<!--   .[, tmp := NULL] -->

<!-- ### merge combinations and zero values to dt_study -->
<!-- dt_study <-  -->
<!--   merge(dt_study, all_combinations, -->
<!--         by = c("Time", "Species", "Genus"), -->
<!--         all = T) %>%  -->
<!--   .[, Abundance := Abundance.x] %>%  -->
<!--   .[is.na(Abundance.x), Abundance := Abundance.y] %>%  -->
<!--   .[, c("Abundance.x", "Abundance.y") := NULL] -->


<!-- # 2)  summarize over multiple entries in the same year -->
<!-- dt_study_genus <- -->
<!--   dt_study[, .(Abundance_sum = sum(Abundance)), -->
<!--     by = c("Time", "Genus")] -->


<!-- # 3)  add relative abundance -->

<!-- ## Species -->
<!-- dt_study[, Abundance_rel := Abundance/ sum(Abundance),  -->
<!--          by = c("Time")] -->

<!-- ## Genus -->
<!-- dt_study_genus[, Abundance_rel := Abundance_sum / sum(Abundance_sum),  -->
<!--          by = c("Time")] -->


<!-- # 3)  Group Species/Genus with relative abundance below a certain threshold as "other" -->

<!-- ## specify threshold for group "other" -->
<!-- threshold = 0.05 -->

<!-- ## add Species "other" for max. rel. abundance < threshold -->
<!-- Species_other <- -->
<!--   dt_study[, .(max_Abundance = max(Abundance_rel)), -->
<!--          by = c("Species")] %>%  -->
<!--   .[max_Abundance < threshold, .(Species)] -->

<!-- dt_study_species <- -->
<!--   dt_study[, Species_grouped := Species] %>%  -->
<!--   .[Species %in% Species_other$Species, Species_grouped := "other"] -->

<!-- ## add Genus "other" for max. rel. abundance < threshold -->
<!-- Genus_other <- -->
<!--   dt_study_genus[, .(max_Abundance = max(Abundance_rel)), -->
<!--          by = c("Genus")] %>%  -->
<!--   .[max_Abundance < threshold, .(Genus)] -->

<!-- dt_study_genus[, Genus_grouped := Genus] %>%  -->
<!--   .[Genus %in% Genus_other$Genus, Genus_grouped := "other"] -->


<!-- # 5)  sum up abundances for group "other" -->

<!-- ## Species -->
<!-- dt_study_species <- -->
<!--   dt_study_species[, .(Abundance_rel = sum(Abundance_rel)), -->
<!--     by = c("Time", "Species_grouped")] -->
<!-- ## Genus -->
<!-- dt_study_genus <- -->
<!--   dt_study_genus[, .(Abundance_sum = sum(Abundance_sum), -->
<!--                      Abundance_rel = sum(Abundance_rel)), -->
<!--     by = c("Time", "Genus_grouped")] -->
<!-- ``` -->



<!-- ### time series plots -->



<!-- ```{r} -->
<!-- ### plot timeseries -->

<!-- # by Species, total -->
<!-- ggplot(dt_study, aes(Time, Abundance)) + -->
<!--   geom_line(aes(col = Species))  + -->
<!--   labs(title = "by Species, total Abundances") + -->
<!--   theme(legend.position = "right") -->

<!-- # by Species, relative -->
<!-- ggplot(dt_study_species, aes(Time, Abundance_rel)) + -->
<!--   geom_bar(aes(fill = Species_grouped), stat = "identity") + -->
<!--   labs(title = "by Species, relative Abundances") + -->
<!--   theme(legend.position = "right") + -->
<!--   scale_fill_viridis(discrete = TRUE, option = "turbo")  -->

<!-- # by Genus, total (Abundance summed up) -->
<!-- ggplot(dt_study_genus, aes(Time, Abundance_sum)) + -->
<!--   geom_line(aes(col = Genus_grouped))  + -->
<!--   labs(title = "by Genus, total Abundances") + -->
<!--   theme(legend.position = "right") + -->
<!--   scale_color_viridis(discrete = TRUE, option = "turbo")  -->

<!-- # by Genus, relative (Abundance summed up) -->
<!-- ggplot(dt_study_genus, aes(Time, Abundance_rel)) + -->
<!--   geom_bar(aes(fill = Genus_grouped), stat = "identity")  + -->
<!--   labs(title = "by Genus, relative Abundances") + -->
<!--   theme(legend.position = "right") + -->
<!--   scale_fill_viridis(discrete = TRUE, option = "turbo")  -->


<!-- # ### rewrite data.table as time series -->
<!-- # dt_study_ts <- -->
<!-- #   dcast(dt_study, Species ~ Time, value.var = "Abundance") -->
<!-- # dt_study_ts[, num_na := rowSums(is.na(dt_study_ts))] -->
<!-- # GEN_SPEC_relevant <- dt_study_ts[num_na < 20]$Species -->
<!-- #  -->
<!-- # dt_study_ts -->
<!-- #  -->
<!-- # dt_study_ts$num_na %>% hist() -->
<!-- #  -->
<!-- # dt_study_ts[is.na(dt_study_ts),] <- 0 -->
<!-- #  -->
<!-- # dt_study_0 <- -->
<!-- #   dt_study_ts[, num_na := NULL] %>%  -->
<!-- #   melt(., id.vars = "Species",  -->
<!-- #        value.name = "Abundance", variable.name = "Time") -->
<!-- #  -->
<!-- #  -->
<!-- # # plot timeseries -->
<!-- # ggplot(dt_study_0, -->
<!-- #        aes(as.numeric(Time) + 1952, Abundance)) + -->
<!-- #   geom_line(aes(col = Species))  + -->
<!-- #   theme(legend.position = "none") -->
<!-- #  -->
<!-- # # plot timeseries -->
<!-- # ggplot(dt_study_0[Species %in% GEN_SPEC_relevant], -->
<!-- #        aes(as.numeric(Time) + 1952, Abundance)) + -->
<!-- #   geom_line(aes(col = Species))  + -->
<!-- #   theme(legend.position = "none") -->

<!-- ``` -->

<br>
## (2)

```{r}
study_id = 414

study_processing_bioTIME(study_id)
```

<!-- #### Information on the study -->

<!-- ```{r, results = "asis", echo = FALSE} -->
<!-- study_Meta <- -->
<!--   dt_biotimeMeta[STUDY_ID == study_id] -->

<!-- table_Meta <- -->
<!--   data.table(Column = colnames(study_Meta), -->
<!--              Info = as.character(study_Meta[1,])) -->

<!-- knitr::kable(table_Meta) -->

<!-- ``` -->

<!-- ### time series plots -->

<!-- ```{r} -->

<!-- # extract data of the specific study -->
<!-- dt_study <- -->
<!--   extract_study(dt_fullquery, ID = study_id) -->

<!-- # summarize over multiple entries in the same year -->
<!-- dt_study_genus <- -->
<!--   dt_study[, .(Abundance_sum = sum(Abundance)), -->
<!--     by = c("Time", "Genus")] -->

<!-- # number of entries per Time and Genus -->
<!-- dt_study[, .N, by= c("Time", "Genus")] -->

<!-- # number of entries per Time and Species -->
<!-- dt_study[, .N, by= c("Time", "Species")] -->


<!-- # plot timeseries -->
<!-- ggplot(dt_study, aes(Time, Abundance)) + -->
<!--   geom_line(aes(col = GENUS_SPECIES))  + -->
<!--   theme(legend.position = "none") -->

<!-- # plot Abundance summed up for same Time -->
<!-- ggplot(dt_study_genus, aes(Time, Abundance_sum)) + -->
<!--   geom_line(aes(col = Genus))  + -->
<!--   theme(legend.position = "bottom") -->


<!-- dt_study_ts <- -->
<!--   dcast(dt_study, GENUS_SPECIES ~ Time, value.var = "Abundance") -->
<!-- dt_study_ts[, num_na := rowSums(is.na(dt_study_ts))] -->
<!-- GEN_SPEC_relevant <- dt_study_ts[num_na < 20]$GENUS_SPECIES -->

<!-- dt_study_ts -->

<!-- dt_study_ts$num_na %>% hist() -->

<!-- dt_study_ts[is.na(dt_study_ts),] <- 0 -->

<!-- dt_study_0 <- -->
<!--   dt_study_ts[, num_na := NULL] %>%  -->
<!--   melt(., id.vars = "GENUS_SPECIES",  -->
<!--        value.name = "Abundance", variable.name = "Time") -->


<!-- # plot timeseries -->
<!-- ggplot(dt_study_0, -->
<!--        aes(as.numeric(Time) + 1952, Abundance)) + -->
<!--   geom_line(aes(col = GENUS_SPECIES))  + -->
<!--   theme(legend.position = "none") -->

<!-- # plot timeseries -->
<!-- ggplot(dt_study_0[GENUS_SPECIES %in% GEN_SPEC_relevant], -->
<!--        aes(as.numeric(Time) + 1952, Abundance)) + -->
<!--   geom_line(aes(col = GENUS_SPECIES))  + -->
<!--   theme(legend.position = "none") -->

<!-- # barplot -->
<!-- ggplot(dt_study, aes(Time, Abundance)) + -->
<!--   geom_bar(aes(fill = Species), stat = "identity", position = "fill") + -->
<!--   theme(legend.position = "none") + -->
<!--   labs(x = "Time") -->



<!-- ``` -->

<br>

## (3)

```{r}
study_id = 478

study_processing_bioTIME(study_id)

```

<!-- #### Information on the study -->

<!-- ```{r, results = "asis", echo = FALSE} -->
<!-- study_Meta <- -->
<!--   dt_biotimeMeta[STUDY_ID == study_id] -->

<!-- table_Meta <- -->
<!--   data.table(Column = colnames(study_Meta), -->
<!--              Info = as.character(study_Meta[1,])) -->

<!-- knitr::kable(table_Meta) -->

<!-- ``` -->

<!-- ### time series plots -->

<!-- ```{r} -->

<!-- # extract data of the specific study -->
<!-- dt_study <- -->
<!--   extract_study(dt_fullquery, ID = study_id) -->

<!-- # summarize over multiple entries in the same year -->
<!-- dt_study_genus <- -->
<!--   dt_study[, .(Abundance_sum = sum(Abundance)), -->
<!--     by = c("Time", "Genus")] -->

<!-- # number of entries per Time and Genus -->
<!-- dt_study[, .N, by= c("Time", "Genus")] -->

<!-- # number of entries per Time and Species -->
<!-- dt_study[, .N, by= c("Time", "Species")] -->


<!-- # plot timeseries -->
<!-- ggplot(dt_study, aes(Time, Abundance)) + -->
<!--   geom_line(aes(col = GENUS_SPECIES))  + -->
<!--   theme(legend.position = "none") -->

<!-- # ggplot(dt_study[GENUS_SPECIES %in% GEN_SPEC_relevant], aes(Time, Abundance)) + -->
<!-- #   geom_line(aes(col = GENUS_SPECIES))  + -->
<!-- #   theme(legend.position = "none") + -->
<!-- #   scale_y_log10() -->

<!-- # plot Abundance summed up for same Time -->
<!-- ggplot(dt_study_genus, aes(Time, Abundance_sum)) + -->
<!--   geom_line(aes(col = Genus))  + -->
<!--   theme(legend.position = "none") -->

<!-- # same plot with y log-scaled -->
<!-- ggplot(dt_study_genus, aes(Time, Abundance_sum)) + -->
<!--   geom_line(aes(col = Genus))  + -->
<!--   theme(legend.position = "bottom") + -->
<!--   scale_y_log10() -->

<!-- dt_study_ts <- -->
<!--   dcast(dt_study, GENUS_SPECIES ~ Time, value.var = "Abundance") -->
<!-- dt_study_ts[, num_na := rowSums(is.na(dt_study_ts))] -->
<!-- GEN_SPEC_relevant <- dt_study_ts[num_na < 10]$GENUS_SPECIES -->

<!-- dt_study_ts$num_na %>% hist() -->

<!-- ``` -->

## (4)

```{r}
study_id = 363

study_processing_bioTIME(study_id)

```


<!-- #### Information on the study -->

<!-- ```{r, results = "asis", echo = FALSE} -->
<!-- study_Meta <- -->
<!--   dt_biotimeMeta[STUDY_ID == study_id] -->

<!-- table_Meta <- -->
<!--   data.table(Column = colnames(study_Meta), -->
<!--              Info = as.character(study_Meta[1,])) -->

<!-- knitr::kable(table_Meta) -->

<!-- ``` -->

<!-- ### time series plots -->

<!-- ```{r} -->

<!-- # extract data of the specific study -->
<!-- dt_study <- -->
<!--   extract_study(dt_fullquery, ID = study_id) -->

<!-- # summarize over multiple entries in the same year -->
<!-- dt_study_genus <- -->
<!--   dt_study[, .(Abundance_sum = sum(Abundance)), -->
<!--     by = c("Time", "Genus")] -->

<!-- # number of entries per Time and Genus -->
<!-- dt_study[, .N, by= c("Time", "Genus")] -->

<!-- # number of entries per Time and Species -->
<!-- dt_study[, .N, by= c("Time", "Species")] -->


<!-- # plot timeseries -->
<!-- ggplot(dt_study, aes(Time, Abundance)) + -->
<!--   geom_line(aes(col = GENUS_SPECIES))  + -->
<!--   theme(legend.position = "none") -->

<!-- # plot timeseries -->
<!-- ggplot(dt_study, aes(Time, Abundance)) + -->
<!--   geom_line(aes(col = GENUS_SPECIES))  + -->
<!--   theme(legend.position = "bottom") -->

<!-- # plot sum by Genus -->
<!-- ggplot(dt_study_genus, aes(Time, Abundance_sum)) + -->
<!--   geom_line(aes(col = Genus))  + -->
<!--   theme(legend.position = "bottom") -->


<!-- dt_study_ts <- -->
<!--   dcast(dt_study_genus, Genus ~ Time, value.var = "Abundance_sum") -->
<!-- dt_study_ts -->

<!-- ``` -->


## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
