---
title: "Time series (Bucci Dataset from cLV paper)"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: github_document
params:
  name: "01b-timeseries-CLVpaper" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "fbad9ade-134a-4a83-a801-9900003f3395")
```

In this document the pre-processing of the "Bucci" dataset from the Compositional Lotka Volterra paper is done.

# Packages

```{r packages, message=FALSE}
library("conflicted")
library(tidyverse)
library(data.table)
library(viridis) # for color palettes
library(phyloseq)
library(microViz)

library(ggfortify) # to autoplot time series
```


```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```


# Bucci dataset

## Read data

```{r}
folderpath <- 
  "C:/Users/Maria/Documents/Masterstudium/Masterarbeit/Literatur/Code/clv/data/bucci/data_cdiff/"
  
#### read the file
dt_bucci_raw <- 
  fread("data/clv/bucci/data_cdiff/counts.txt", header = T)

knitr::kable(head(dt_bucci_raw))

# read metadata file
dt_bucci_meta <-
  fread("data/clv/bucci/data_cdiff/metadata.txt", header = T) %>% 
  .[, Time := measurementid] %>% 
  .[, Time_factor := ordered(Time)] %>% 
  .[, measurementid := NULL]

head(dt_bucci_meta)

# read taxonomic information
dt_tax_bucci <-
  fread(paste0(folderpath, "taxonomy_table.csv"), header = T)

```

## Info 

... on the time points: When was the data collected?

Fecal pellets were collected at days 0.75, 1, 2, 3, 4, 6, 8, 10, 14, 17, 21, 24, and 28 of the initial colonization 
and at days 0.75, 1, 2, 3, 4, 6, 8, 10, 14, 17, 21, 24, and 28 post-infection with C. difficile.

--> Time (prev. measurementid)


## Phyloseq

### Prepare otu table

```{r}
mat_otu_bucci <-
  dt_bucci_raw %>%
    tibble::column_to_rownames("Species")
colnames(mat_otu_bucci) <-
  sprintf("ID-%03d", as.numeric(colnames(mat_otu_bucci)))

OTU_bucci = otu_table(mat_otu_bucci, taxa_are_rows = TRUE)

```


### Prepare sample table (metadata)

```{r}
mat_samples_bucci <-
  copy(dt_bucci_meta) %>% 
  .[, sampleID := sprintf("ID-%03d", sampleID)] %>%
  tibble::column_to_rownames("sampleID") 

samples_bucci = sample_data(mat_samples_bucci)

```


### Generate taxonomic table

```{r}
mat_tax_bucci <-
  dt_tax_bucci %>% 
  tibble::column_to_rownames("Names") 

TAX_bucci = tax_table(as.matrix(mat_tax_bucci))

```


### Create phyloseq object

```{r}
ps_bucci_all <- phyloseq(OTU_bucci, TAX_bucci, samples_bucci)
```


## Transform counts to relative counts

```{r}
# change count data to relative counts
# filter for most abundant species

ps_bucci_rel <- 
  ps_bucci_all %>% 
  transform_sample_counts(function(x) x / sum(x))

```


### Devide dataset by subjects

```{r}
# get a subset for each subject 

ps_bucci_1 <-
  subset_samples(ps_bucci_rel, subjectID == 1)

ps_bucci_2 <-
  subset_samples(ps_bucci_rel, subjectID == 2)

ps_bucci_3 <-
  subset_samples(ps_bucci_rel, subjectID == 3)

ps_bucci_4 <-
  subset_samples(ps_bucci_rel, subjectID == 4)

ps_bucci_5 <-
  subset_samples(ps_bucci_rel, subjectID == 5)

```


### Get one dataset, that contains mean of all subjects in one

```{r}
ps_bucci_mean <-
  merge_samples(ps_bucci_all, "Time") %>% 
  transform_sample_counts(function(x) x / sum(x))
```



### Overview of datasets (and their taxonimic levels)

```{r}
ps_bucci_1
ps_bucci_2
ps_bucci_3
ps_bucci_4
ps_bucci_5

# # summarize how many unique levels are available for each taxonomic rank
# # (for overview table)
# for(subject in c(seq(5), "all")){
#   cat("----------------------------------")
#   cat("\n")
#   cat("Subject ", subject, "\n")
#   cat("----------------------------------\n")
#   for(tax_rank in c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")) {
#     cat(tax_rank, ": \t")
#     cat(" uniqueN:   ", tax_table(get(paste0("ps_bucci_", subject)))[, tax_rank] %>% uniqueN(), "\n")
#     cat("\t\t NAs:       ", (sum(is.na(tax_table(get(paste0("ps_bucci_", subject)))[, tax_rank]))), "\n")
#   }
#   cat("\n")
# }
```


### plot data

```{r}
# plot for all subjects
plot_bar(ps_bucci_mean,
         x = "Time_factor",
         fill = "Species",
         title = "All Subjects") +
  geom_bar(aes(color=Species, fill=Species), stat="identity", position="stack")

```


```{r}
# plots for each subject separately
for (id in 1:5) {
  plt_tmp <-
    plot_bar(get(paste0("ps_bucci_", id)),
             x = "Time_factor",
             fill = "Species",
             title = paste("Subject", id)) +
    geom_bar(aes(color=Species, fill=Species), stat="identity", position="stack")
  
  print(plt_tmp)
}
```


## Save Phyloseq Objects

```{r}
for (id in c(1:5, "mean")) {
  saveRDS(get(paste0("ps_bucci_", id)),
          path_target(paste0("ps_bucci_subject_", id, "_rel_count.rds")))
}
```


## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
