---
title: "Simulation of synthetic time series with second order interactions (miaSim)"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: github_document
params:
  name: "01e-timeseries-miaSim" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "097d888c-3dd7-4302-ad02-3bed36ed3cfe")
```

The purpose of this document is to simulate some synthetic time series datasets with different numbers of species and some 2nd order interactions (using the miaSim package).
These can then be used for testing the DeePyMoD algorithm.

# Packages

```{r packages, message=FALSE}
library("conflicted")
library(tidyverse)
library(data.table)

# BiocManager::install("miaSim")
library(miaSim)
library(miaViz)

library(deSolve)
library(R.matlab)

library(ggfortify) # to autoplot time series
```


```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```


# miaSim Shiny App

```{r}
# library(miaSimShiny)
# ## run the shiny app for miaSim
# miaSimShiny::run_app()
```



# miaSim manual Simulation

## Function to simulate, save and plot GLV

```{r}
# simulates generalized Lotka Volterra with miaSim (dx/dt = x(b + Ax))
# for t = 1,...,500
# saves the glv time series as .mat file
# plots the glv time series
# sim_name: give the simulation a name, e.g. "_oscillating"

save_and_plot_glv <-
  function(n_species, A_matrix, growth_rates = NULL, x0 = NULL, sim_name = "") {
    if(is.null(growth_rates)){
      growth_rates <- runif(n_species)
    }
    if(is.null(x0)){
      x0 <- runif(n_species)
    }
      
    # simulate GLV
    glv_tmp <-
      simulateGLV(
        n_species = n_species,
        A = A_matrix,
        x = x0,
        b = growth_rates,
        t_start = 0,
        t_store = 500
      )
    
    # write .mat file for python
    glv_tmp_mat <- data.matrix(cbind(1:500, t(glv_tmp)))
    writeMat(paste0(path_target(),"/miaSim_GLV_", n_species, "species",
                    sim_name, ".mat"),
             ts_glv = glv_tmp_mat)
    
    autoplot(ts(glv_tmp_mat[, 1:n_species+1]), facets = F)
  }

```


## Simple Examples for simulating GLV with 2 Species

### a)

```{r}
# Generate random interaction matrix for GLV (2 Species)
# A_matrix <- randomA(n_species = 2)
A_matrix <- matrix(c(-1, 4/3, 1, -1), nrow = 2)
A_matrix
growth_rates <- c(2/3, 1)

A_matrix <- matrix(c(-0.5, 0, 0.1, -0.5), nrow = 2)
# A_matrix
# growth_rates <- c(0.5, 0.3)

save_and_plot_glv(A_matrix, growth_rates, 
                  n_species = 2, sim_name = "_V1")

```

### b)

```{r}
# Generate random interaction matrix for GLV (2 Species)
A_matrix <- matrix(c(-0.5, 2, 0, -0.5), nrow = 2)
A_matrix

growth_rates = c(1, 2)

set.seed(321)


save_and_plot_glv(n_species = 2, A_matrix, growth_rates, 
                  sim_name = "_V3")

```


## General gLV for arbitrary number n of species

### a) 5 Species

```{r}
n = 5

# Generate random interaction matrix for GLV (3 Species)
set.seed(246)
A_matrix <- randomA(n_species = n)
A_matrix <- matrix(c(-0.5, 20, 0, 0, 0,
                     -28, -0.5, -3, 0, 0,
                     0, 10, -0.5, 0, 0,
                     -3, 0, 0, -0.5, 7.5,
                     5, 0, 0, -4.4, -0.5), nrow = n)
A_matrix

# specify growth rates or use rep(1, n) by default
growth_rates = c(10, -1, 1, 1, -1)
growth_rates

save_and_plot_glv(n_species = n, A_matrix, growth_rates, 
                  sim_name = "_V1")
```

### b) 2 species

```{r}
n = 2

# Generate random interaction matrix for GLV (3 Species)
set.seed(246)
A_matrix <- randomA(n_species = n)
A_matrix <- matrix(c(-0.5, 15,
                     -30, -0.5), nrow = n)
A_matrix

# specify growth rates or use rep(1, n) by default
growth_rates = c(2/3, -1)

save_and_plot_glv(n_species = n, A_matrix, growth_rates, 
                  sim_name = "_oscillating")
```


### c) 3 Species

```{r}
n = 3

# Generate random interaction matrix for GLV (3 Species)
set.seed(246)
A_matrix <- randomA(n_species = n)
A_matrix <- matrix(c(-0.5, 5, 2,
                     -3, -0.5, 3,
                     0, 0, -0.5), nrow = n)
A_matrix

# specify growth rates or use rep(1, n) by default
growth_rates = c(2/3, -1, -0.7)


save_and_plot_glv(n_species = n, A_matrix, growth_rates, 
                  sim_name = "_oscillating")
```


### c.1) 4 Species

```{r}
n = 4

# Generate random interaction matrix for GLV
A_matrix <- matrix(c(-0.5, 80, 0, 0,
                     -50, -0.5, 3, 0,
                     0, 0, -0.5, 20,
                     0, 0, -10, -0.5), nrow = n)
A_matrix

# specify growth rates or use rep(1, n) by default
growth_rates = c(5, -10, 7, -2)


save_and_plot_glv(n_species = n, A_matrix, growth_rates, 
                  sim_name = "_oscillating_three")
```


### c.2) 4 Species

```{r}
n = 4

# Generate random interaction matrix for GLV
A_matrix <- matrix(c(-0.5, 80, 0, 0,
                     -50, -0.5, 0, 0,
                     0, 0, -0.5, 20,
                     0, 0, -10, -0.5), nrow = n)
A_matrix

# specify growth rates or use rep(1, n) by default
growth_rates = c(5, -10, 7, -2)


save_and_plot_glv(n_species = n, A_matrix, growth_rates, 
                  sim_name = "_oscillating_zero")

```


### d) 4 Species

```{r}
n = 4

# Generate random interaction matrix for GLV
A_matrix <- matrix(c(-0.5, 30, 0, 0,
                     0, -0.5, -15, 0,
                     0, 0, -0.5, 20,
                     -10, 0, 0, -0.5), nrow = n)
A_matrix

# specify growth rates or use rep(1, n) by default
growth_rates = c(5, -10, 7, 0.5)

x0 = c(5,5,5,5)

save_and_plot_glv(n_species = n, A_matrix, growth_rates, x0, 
                  sim_name = "_Vd")
```

## Example with 4 species

```{r}
n = 4

# Interaction matrix with Power-Law network adjacency matrix (from tutorial)
set.seed(321)
A_normal <- powerlawA(n_species = n, alpha = 3)
A_normal

# simulate GLV
set.seed(123)
save_and_plot_glv(n_species = n, A_matrix = A_normal)
```

```{r}
A <- miaSim::powerlawA(4, alpha = 1.01)
A

save_and_plot_glv(n_species = 4, A_matrix = A)
```



## Large Example for simulating GLV 

### a) with 30 Species

```{r}
n = 30

# Generate random interaction matrix for GLV (2 Species)
set.seed(12)
A_matrix <- randomA(n_species = 30)
A_matrix

# simulate GLV
save_and_plot_glv(n_species = n, A_matrix)

```


### b) with 10 Species

```{r}
n = 10

# Generate random interaction matrix for GLV (2 Species)
# set.seed(2)
# A_matrix <- randomA(n_species = n)
A_matrix <- matrix(c(-0.5, 30, 0, 0, 0, 0, 0, 0, 0, 0,
                     0, -0.5, 0, 0, 0, 0, 0, 0, 0, 0,
                     0, 0, -0.5, 0, 0, 0, 0, 0, 0, 0,
                     0, 0, 0, -0.5, 0, 0, 12, 0, 0, 0,
                     0, 0, 0, 0, -0.5, 0, 0, 0, 0, 0,
                     0, 0, 0, 0, 0, -0.5, 0, 0, 0, 0,
                     0, 0, -15, 0, 0, 0, -0.5, 0, 0, 0,
                     0, 0, 0, 0, 0, 0, 0, -0.5, 0, 0,
                     0, 0, 0, 0, 0, 0, 0, 0, -0.5, 0,
                     0, 0, 0, 0, 0, 0, 0, 0, 2, -0.5),
                   nrow = n)
A_matrix

growth_rates = rep(1, n)
growth_rates

save_and_plot_glv(n_species = n, A_matrix, growth_rates)

```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
